Density-dependent selection and the population ecology of relative fitness

### 2017-11-28 manuscript.tex
l

The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Other than densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models. The incompatibility is particularly apparent in some of the most widely-used models of density-regulated population growth (to illustrate our point we stick to the simple haploid case): 1) The logistic model: the type with the greatest saturation population density excludes the others. 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Except in special cases, differences in $R^*$ entails differences in saturation population density. 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH). Thus, it would seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

It is not even clear how to connect relative fitness models to density-regulated growth, since they only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.

@@

The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation of a more complicated long-term eco-evolutionary process \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate dramatically over time, and coupled to this, that a population's density can vary by orders of magnitude over a few generations as a routine feature of its ecology. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}. Moreover, the short-term approximation automatically precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction. Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.

@@

--- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a)

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; $K_{12}$ and $K_{21}$ are related to competition between types. For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a). Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Blah}
\end{figure}

In Fig.~\ref{fig:LVvslottery} we show a phase diagram for Eq.~\eqref{eq:master}

>>>>

The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Apart from the type densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models. Some of the most widely-used models of density-regulated population growth are indeed not compatible: 1) In the logistic model, the type with the greatest saturation population density excludes the others. 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Differences in $R^*$ will often entail differences in saturation density. 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH). It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments. It is not even clear how relative fitness models connect to general ecological models of the form Eq.~\eqref{eq:macgeneral}, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.

@@

The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation \cite[pp. 277]{ewens_2004}. That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences. Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction. Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology. These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}. The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}. Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction. Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.

@@

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, behaviour which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; selection on these saturation densities is commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density. MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} (a) Lotka-Volt}
\end{figure}

Fig.~\ref{fig:LVvslottery} shows two examples of selection when both of two types have the same saturation density $N=K_{11}=K_{22}$. The first shows exlusion by a superior Lotka-Volterra competitor. Note that the trajectories do not follow a line of constant  $N$. The reason for this is that inter-type competition must be stronger than intra-type competition for exclusion to occur, which causes a dip in $N$ over a selective sweep (except for the special case  .

$c$-selection in the density-dependent lottery model, Eq.~\eqref{eq:master}.
